% =============================================================================
% Chapter 2: Electromagnetic Theory of a Uniaxial Medium
% =============================================================================

\chapter{Electromagnetic Theory of a Uniaxial Medium}
\label{ch:em_theory}

This chapter builds the electromagnetic foundation that everything else in this
report rests on. We begin with the simplest possible case, a monochromatic
plane wave in a homogeneous medium, and work our way step by step toward the
full dispersion relation of a uniaxial anisotropic material. We then introduce
material loss and show rigorously, following Jackson et al.\
(2024)~\cite{jackson2024}, how it transforms the open hyperbolic
isofrequency contour of an idealised lossless medium into a closed,
loss-broadened surface.

% -----------------------------------------------------------------------------
\section{The Plane Wave and the Complex Refractive Index}
% -----------------------------------------------------------------------------

All the phenomena discussed in this report, reflection, transmission,
birefringence, and dispersion, are consequences of how electromagnetic waves
interact with matter. The simplest wave to study is a monochromatic plane wave,
in which the electric field oscillates everywhere with the same frequency
$\omega$ and varies spatially only through a complex exponential. We write it as
\begin{equation}
  \Evec(\bm{r},t) = \Evec_0 \, \ee^{\ii(\kvec \cdot \bm{r} - \omega t)},
  \label{eq:ch2_planewave}
\end{equation}
where $\Evec_0$ is a constant amplitude vector, $\kvec$ is the wavevector
pointing in the direction of propagation, and $\bm{r}$ is the position vector.

We use the \textbf{physics convention} $\ee^{-\ii\omega t}$ for the time
dependence throughout this report. This choice determines the sign of the
imaginary part of all complex optical quantities. It is worth stating this
explicitly because the opposite convention $\ee^{+\ii\omega t}$ (common in
engineering texts and Hecht's \emph{Optics}~\cite{hecht}) leads to conjugate
signs everywhere. When copying a formula from a source using the opposite
convention, the imaginary parts must be negated.

\begin{warnbox}
\textbf{Convention.} This report uses $\ee^{-\ii\omega t}$. The complex
refractive index is $\nt = n + \ii\kappa$ with $n, \kappa \geq 0$. The complex
permittivity is $\tilde{\eps} = \nt^2 = (n^2 - \kappa^2) + \ii\,2n\kappa$,
with a positive imaginary part for a passive absorbing material. Hecht uses
$e^{+\ii\omega t}$, giving $\nt = n - \ii\kappa$, which is the complex
conjugate of ours. Flip the sign of all imaginary parts when borrowing from
Hecht.
\end{warnbox}

In a real material, part of the electromagnetic energy is converted to heat.
The imaginary part $\kappa$ of the refractive index quantifies this absorption.
To see how, substitute $\kvec = (\omega/c)\nt\,\hat{\bm{k}}$ for a wave
travelling along $+z$ and expand the exponential:
\begin{equation}
  E = E_0 \,
  \underbrace{\ee^{-\frac{\omega\kappa}{c}\,z}}_{\text{amplitude decay}}
  \underbrace{\ee^{\ii\!\left(\frac{\omega n}{c}\,z - \omega t\right)}}_{\text{oscillation}}.
  \label{eq:ch2_absorbing_wave}
\end{equation}
The phase velocity of the oscillation is $c/n$, so $n$ sets how fast the
wavefronts advance. The amplitude decays over a characteristic length
$L_p = c/(\omega\kappa)$, called the propagation length or skin depth. For
nickel at $700$\,nm wavelength, $\kappa \approx 4.7$~\cite{johnson_christy},
giving $L_p \approx 12$\,nm. This is why Ni films of only $5$\,nm thickness
already absorb roughly half the light passing through them.

The relationship between the refractive index and the permittivity in a
non-magnetic material ($\mu_r = 1$) is simply
\begin{equation}
  \tilde{\eps} = \nt^2 = (n^2 - \kappa^2) + \ii\,2n\kappa.
  \label{eq:ch2_eps_nk}
\end{equation}
For a metal at visible wavelengths, $n \sim \kappa$, which means
$\Re(\tilde{\eps}) = n^2 - \kappa^2$ can be negative while $\Im(\tilde{\eps})
= 2n\kappa$ is always large and positive. For nickel at $700$\,nm:
$n = 2.12$, $\kappa = 4.70$~\cite{johnson_christy}, giving
$\Re(\eps_\text{Ni}) \approx 2.12^2 - 4.70^2 = 4.49 - 22.09 = -17.6$ and
$\Im(\eps_\text{Ni}) \approx 2 \times 2.12 \times 4.70 = 19.9$.
The large negative real part and large imaginary part together make Ni a
strongly metallic, highly lossy material.

% -----------------------------------------------------------------------------
\section{Geometry of the Problem: Interfaces and Wavevectors}
% -----------------------------------------------------------------------------

The sample in this report is a planar multilayer stack whose interfaces are
all perpendicular to the $z$ axis. Figure~\ref{fig:ch2_geometry} shows the
relevant geometry. A plane wave arrives from above (the air side) at an
angle $\theta$ measured from the normal to the interfaces. The plane of
incidence, the plane containing both the incident beam and the normal, is
the $x$--$z$ plane.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.1,
  axis/.style={-{Latex[length=2mm]}, thick},
  wave/.style={-{Latex[length=2.5mm]}, very thick, color=KITblue},
  comp/.style={-{Latex[length=2mm]}, thick, color=coral},
  ifc/.style={thick, color=navy}
]

% Interface
\fill[gray!10] (-3.5,-2.2) rectangle (3.5,0);
\draw[ifc, line width=1.2pt] (-3.5,0) -- (3.5,0);
\node[right, font=\small, color=navy] at (3.5, 0) {interface};

% Medium labels
\node[font=\small] at (-2.8, 1.0) {Medium 1 (air)};
\node[font=\small, color=navy!70] at (-2.8,-0.6) {$n_1 = 1$};
\node[font=\small] at (-2.8,-1.2) {Medium 2 (material)};
\node[font=\small, color=navy!70] at (-2.8,-1.8) {$\nt = n + \ii\kappa$};

% Normal axis
\draw[axis, color=navy!50, dashed] (0,-2.2) -- (0,2.2)
  node[above, font=\small] {normal ($z$)};

% Incident k vector
\draw[wave] (-1.8, 1.8) -- (0,0)
  node[midway, above left, font=\small, color=KITblue] {$\kvec$};

% Angle theta
\draw[color=navy] (0,0.7) arc (90:135:0.7)
  node[midway, above, font=\small] {$\theta$};

% kx component (horizontal arrow along interface)
\draw[comp] (0,0) -- (-1.8,0)
  node[midway, below, font=\small, color=coral] {$k_x = \frac{\omega}{c}\sin\theta$};

% kz component (vertical arrow downward into material)
\draw[comp] (0,0) -- (0,-1.8)
  node[midway, right, font=\small, color=coral] {$k_z$};

% Show kx conserved with dotted line
\draw[dashed, color=coral!60] (-1.8, 0) -- (-1.8, -1.4);
\draw[comp, color=coral!60] (0,0) -- (-1.8,-1.4);
\node[below left, font=\scriptsize, color=coral!80] at (-1.8,-1.4)
  {$k_z'$ (refracted)};
\node[font=\scriptsize, color=coral] at (-0.9, -1.4)
  {$k_x$ conserved};

% Coordinate label
\draw[axis, color=navy!40] (2.5, -2.0) -- (3.2,-2.0)
  node[right, font=\scriptsize] {$x$};
\draw[axis, color=navy!40] (2.5, -2.0) -- (2.5,-1.3)
  node[above, font=\scriptsize] {$z$};

\end{tikzpicture}
\caption{Geometry of a plane wave incident on a planar interface between
  air (medium 1, above) and an absorbing material (medium 2, below). The
  interface is normal to $z$. The wavevector $\kvec$ lies in the $x$--$z$
  plane at angle $\theta$ from the normal. The boundary condition at the
  interface forces the tangential (in-plane) component $k_x$ to be the same
  in both media, while the normal component $k_z$ changes to satisfy the
  dispersion relation in each medium.}
\label{fig:ch2_geometry}
\end{figure}

The wavevector in any medium must satisfy the dispersion relation, which comes
directly from Maxwell's equations:
\begin{equation}
  k_x^2 + k_z^2 = \left(\frac{\omega\,\nt}{c}\right)^2 = \tilde{\eps}\,k_0^2,
  \qquad k_0 = \frac{\omega}{c} = \frac{2\pi}{\lambda}.
  \label{eq:ch2_dispersion}
\end{equation}
Here $k_0$ is the free-space wavenumber. When the wave crosses an interface,
the fields must match at every point along the boundary at every instant. This
matching condition requires the phase of the wave to be the same on both sides
of the interface at all times. Because the phase is $k_x x - \omega t$ along
the surface, the only way to satisfy this for all $x$ and $t$ simultaneously is
if $k_x$ is identical in both media. This is the fundamental reason for
Snell's law, stated here as a wavevector conservation:
\begin{equation}
  \boxed{k_x = \frac{\omega}{c}\,\sin\theta = \text{constant across all interfaces.}}
  \label{eq:ch2_kx_conserved}
\end{equation}
Once $k_x$ is fixed by the angle of incidence, the normal component $k_z$ in
each medium follows from the dispersion relation:
\begin{equation}
  k_z = \sqrt{\tilde{\eps}\,k_0^2 - k_x^2}.
  \label{eq:ch2_kz}
\end{equation}
For an absorbing medium, $\tilde{\eps}$ is complex, so $k_z$ is generally
complex even when $k_x$ is real. Working with $k_z$ directly avoids the
awkward complex arcsine that would appear in a refraction-angle formulation.

% -----------------------------------------------------------------------------
\section{Fresnel Coefficients at a Single Interface}
% -----------------------------------------------------------------------------

When a plane wave meets a planar interface, it splits into a reflected wave and
a transmitted wave. The amplitudes of these waves relative to the incident wave
are given by the Fresnel coefficients, derived by applying the boundary
conditions to both field components simultaneously~\cite{hecht}.

For the \textbf{TE (s-polarised)} wave, the electric field is perpendicular to
the plane of incidence ($E$ points along $\hat{y}$). The boundary conditions
require continuity of $E_y$ and of the tangential magnetic field $H_x$. Using
$k_z$ rather than angles, the reflection coefficient at an interface from
medium 1 to medium 2 is
\begin{equation}
  r_{12}^{\,s} = \frac{k_{z1} - k_{z2}}{k_{z1} + k_{z2}}.
  \label{eq:ch2_rs}
\end{equation}

For the \textbf{TM (p-polarised)} wave, the electric field lies in the plane
of incidence ($E$ has components along both $\hat{x}$ and $\hat{z}$), while
the magnetic field is along $\hat{y}$. The boundary conditions require
continuity of $H_y$ and of the tangential electric field $E_x$. In an
isotropic medium with permittivity $\tilde{\eps}$, the result is
\begin{equation}
  r_{12}^{\,p} = \frac{\tilde{\eps}_2\,k_{z1} - \tilde{\eps}_1\,k_{z2}}
                      {\tilde{\eps}_2\,k_{z1} + \tilde{\eps}_1\,k_{z2}}.
  \label{eq:ch2_rp}
\end{equation}
By the Stokes relations, the reflection coefficient for the reverse direction
is $r_{21} = -r_{12}$, and the transmission amplitudes satisfy
$t_{12}t_{21} = 1 - r_{12}^2$. These relations will be useful in the
multiple-beam summation below.

The power reflectance is $R = |r|^2$, which is always real and between 0
and 1. The transmittance involves a ratio of the normal power fluxes and
takes the form $T = (\Re(k_{z2})/k_{z1})\,|t|^2$ for a non-absorbing
second medium. For our multilayer problem, we will compute $R$ and $T$
through the transfer matrix formalism of Chapter~\ref{ch:tmm_derivation},
which handles all of this automatically.

% -----------------------------------------------------------------------------
\section{Multiple Reflections in a Slab: the Airy Formulae}
% -----------------------------------------------------------------------------

A single thin film sitting between two semi-infinite media supports an
infinite sequence of reflections. Each successive internal reflection
contributes a partial wave to the total reflected and transmitted fields.
Summing this infinite geometric series gives the Airy amplitude
formulae~\cite{hecht}.

Consider a slab of thickness $d$ and complex refractive index $\nt$, with
air on both sides. Each round trip inside the slab accumulates a phase
$2\beta$, where the one-way phase is
\begin{equation}
  \beta = k_z\,d = k_0\,d\,\sqrt{\tilde{\eps} - \sin^2\theta}.
  \label{eq:ch2_beta}
\end{equation}
The total reflected amplitude is the sum of the directly reflected wave plus
all waves that escape after one, two, three, ... round trips:
\begin{equation}
  r_\text{slab} = r_{12} + t_{12}\,t_{21}\,r_{23}\,\ee^{2\ii\beta}
    \sum_{m=0}^{\infty}(r_{21}\,r_{23}\,\ee^{2\ii\beta})^m
  = \frac{r_{12}(1 - \ee^{2\ii\beta})}{1 - r_{12}^2\,\ee^{2\ii\beta}},
  \label{eq:ch2_rslab}
\end{equation}
where the second equality uses the fact that $r_{23} = -r_{21} = r_{12}$
for a symmetric slab (same medium on both sides) and sums the geometric
series. The transmitted amplitude is similarly
\begin{equation}
  t_\text{slab} = \frac{(1 - r_{12}^2)\,\ee^{\ii\beta}}{1 - r_{12}^2\,\ee^{2\ii\beta}}.
  \label{eq:ch2_tslab}
\end{equation}
The measurable power coefficients are then
\begin{equation}
  R = |r_\text{slab}|^2, \qquad T = |t_\text{slab}|^2.
  \label{eq:ch2_RT}
\end{equation}
For an absorbing slab, part of the incident power is converted to heat, so
$R + T < 1$. The absorbed fraction $A = 1 - R - T$ is the absorptance.

\begin{resultbox}
\textbf{Why no Fabry-P\'erot fringes are visible in this sample.}
The Airy formula predicts oscillations in $T(\lambda)$ whenever the round-trip
phase satisfies $2\,\Re(\beta) = 2m\pi$ for integer $m$. These are the
Fabry-P\'erot fringes. For the $44$\,nm Ni/TiO$_2$ stack at $700$\,nm, the
round-trip phase is $2k_0 d\,\Re(\sqrt{\tilde\eps}) \approx 2 \times 0.40 \times
\Re(\sqrt{\eps_\text{eff}}) \approx 0.8\,\text{rad}$, far less than $2\pi$.
The first fringe would appear near $200$\,nm wavelength, deep in the ultraviolet.
The measurement window ($550$ to $1000$\,nm) contains no fringes. This rules
out the direct Fabry-P\'erot reconstruction of the wavevector and forces us
onto the model-based inverse approach described in
Chapters~\ref{ch:why_rt_failed} and~\ref{ch:inverse_methodology}.
\end{resultbox}

% -----------------------------------------------------------------------------
\section{The Uniaxial Permittivity Tensor}
% -----------------------------------------------------------------------------

An isotropic medium responds the same way to an electric field pointing in
any direction, so a single complex number $\tilde\eps$ is sufficient to
describe its optical response. An anisotropic medium responds differently
depending on the direction of the applied field. For a layered metal-dielectric
stack, this anisotropy arises from the geometry: fields oscillating within
the layer planes experience a different effective environment from fields
oriented perpendicular to the layers.

The most general linear relationship between the displacement field
$\Dvec$ and the electric field $\Evec$ in such a medium is
$\Dvec = \eps_0\,\bar{\bar\eps}\cdot\Evec$, where $\bar{\bar\eps}$ is the
permittivity tensor. For a multilayer stack whose layers are perpendicular to
$z$, the two independent directions are in-plane ($x$ and $y$) and
out-of-plane ($z$). By the rotational symmetry of the stack in the $xy$ plane,
the tensor is diagonal with two distinct values:
\begin{equation}
  \bar{\bar\eps} = \begin{pmatrix}
    \epar & 0 & 0 \\ 0 & \epar & 0 \\ 0 & 0 & \eperp
  \end{pmatrix}.
  \label{eq:ch2_tensor}
\end{equation}
Here $\epar \equiv \eps_x = \eps_y$ is the in-plane permittivity, governing
the response to electric fields oscillating parallel to the layer planes, and
$\eperp \equiv \eps_z$ is the out-of-plane permittivity, governing the
response to fields oriented along the growth direction. Both are complex
numbers. Their values for the Ni/TiO$_2$ stack are derived from the
constituent materials in Chapter~\ref{ch:emt} and extracted from ellipsometry
in Chapter~\ref{ch:inverse_methodology}.

A medium described by this diagonal tensor is called
\textbf{uniaxial}, because it has a single special axis (the $z$ axis,
called the optic axis) along which the optical response differs from all
perpendicular directions. Natural uniaxial crystals such as calcite and
quartz have $\epar$ and $\eperp$ both real and positive. A
metal-dielectric multilayer can, under the right conditions, have $\epar$
and $\eperp$ with opposite signs, as we now show.

% -----------------------------------------------------------------------------
\section{The Wave Equation in a Uniaxial Medium}
% -----------------------------------------------------------------------------

With the permittivity tensor in hand, we can now derive the dispersion
relation for electromagnetic waves propagating in such a medium. The
derivation starts from Maxwell's curl equations in a source-free,
non-magnetic medium:
\begin{align}
  \nabla \times \Evec &= \ii\omega\mu_0\,\Hvec,
  \label{eq:ch2_faraday} \\
  \nabla \times \Hvec &= -\ii\omega\eps_0\,\bar{\bar\eps}\cdot\Evec.
  \label{eq:ch2_ampere}
\end{align}
For a plane wave, all spatial derivatives act as $\nabla \to \ii\kvec$, so
the curl operations become cross products. Applying $\nabla\times$ to
equation~\eqref{eq:ch2_faraday} and substituting~\eqref{eq:ch2_ampere}
yields the vector wave equation:
\begin{equation}
  \kvec \times (\kvec \times \Evec) = -k_0^2\,\bar{\bar\eps}\cdot\Evec.
  \label{eq:ch2_wave_eq_raw}
\end{equation}
Using the vector identity $\kvec \times (\kvec \times \Evec) =
\kvec(\kvec\cdot\Evec) - k^2\Evec$ (where $k^2 = \kvec\cdot\kvec$),
this becomes
\begin{equation}
  \kvec(\kvec\cdot\Evec) - k^2\Evec = -k_0^2\,\bar{\bar\eps}\cdot\Evec.
  \label{eq:ch2_master}
\end{equation}
This is the master equation for electromagnetic wave propagation in an
anisotropic medium. Everything that follows is the linear algebra of this
equation.

% -----------------------------------------------------------------------------
\section{The TM Dispersion Relation}
% -----------------------------------------------------------------------------

We place $\kvec$ in the $x$--$z$ plane (by rotational symmetry, this
loses no generality): $\kvec = k_x\,\hat{x} + k_z\,\hat{z}$, where
$k_x = k_0\sin\theta$ is fixed by the angle of incidence and
$k_z$ is what we want to find.

For the \textbf{TM (p-polarised)} mode, the electric field lies in the plane
of incidence: $\Evec = (E_x, 0, E_z)$. The magnetic field is $\Hvec\parallel\hat{y}$.
Computing the dot product and cross products in~\eqref{eq:ch2_master} and
reading off the $x$ and $z$ components gives the homogeneous $2\times 2$
system:
\begin{equation}
  \begin{pmatrix}
    k_0^2\epar - k_z^2 & k_x k_z \\
    k_x k_z & k_0^2\eperp - k_x^2
  \end{pmatrix}
  \begin{pmatrix} E_x \\ E_z \end{pmatrix}
  = \begin{pmatrix} 0 \\ 0 \end{pmatrix}.
  \label{eq:ch2_2x2}
\end{equation}
A non-trivial electric field $(E_x, E_z) \neq (0,0)$ exists only if the
determinant of this matrix is zero:
\begin{equation}
  (k_0^2\epar - k_z^2)(k_0^2\eperp - k_x^2) - k_x^2 k_z^2 = 0.
  \label{eq:ch2_det}
\end{equation}
Expanding and cancelling the cross term $k_x^2 k_z^2$ gives
$k_0^4\epar\eperp - k_0^2\epar k_x^2 - k_0^2\eperp k_z^2 = 0$.
Dividing by $k_0^2\epar\eperp$ yields the \textbf{TM dispersion relation}:
\begin{equation}
  \boxed{\frac{k_x^2}{\eperp} + \frac{k_z^2}{\epar} = k_0^2.}
  \label{eq:ch2_TM_disp}
\end{equation}
This is the isofrequency equation: at fixed frequency $\omega$ (and hence
fixed $k_0$), it defines the curve in $(k_x, k_z)$ space of all allowed
propagation directions. Its shape depends entirely on the signs of
$\Re(\epar)$ and $\Re(\eperp)$.

Solving equation~\eqref{eq:ch2_TM_disp} for $k_z$ gives
\begin{equation}
  k_z = \sqrt{\epar\!\left(k_0^2 - \frac{k_x^2}{\eperp}\right)}
      = k_0\sqrt{\epar - \frac{\epar}{\eperp}\sin^2\theta},
  \label{eq:ch2_kz_aniso}
\end{equation}
which is the $k_z$ that appears in the anisotropic transfer matrix
(Chapter~\ref{ch:tmm_derivation}). When $\epar = \eperp = \tilde\eps$,
this reduces to $k_z = k_0\sqrt{\tilde\eps - \sin^2\theta}$, recovering
the isotropic result.

% -----------------------------------------------------------------------------
\section{Why TE Polarisation Cannot Be Hyperbolic}
\label{sec:ch2_TE}
% -----------------------------------------------------------------------------

The TM relation~\eqref{eq:ch2_TM_disp} involves both $\epar$ and $\eperp$
and can be hyperbolic. The TE polarisation behaves very differently.

For the \textbf{TE (s-polarised)} mode, the electric field is $\Evec = E_y\hat{y}$,
perpendicular to the plane of incidence. The dot product $\kvec\cdot\Evec$
vanishes because $\kvec$ has no $y$ component. The master
equation~\eqref{eq:ch2_master} therefore reduces to
\begin{equation}
  -k^2\Evec = -k_0^2\,\bar{\bar\eps}\cdot\Evec.
  \label{eq:ch2_TE_wave}
\end{equation}
The tensor acts on $E_y\hat{y}$ to give $\bar{\bar\eps}\cdot(E_y\hat{y}) =
\epar E_y\hat{y}$, because the $yy$ component of the tensor is $\epar$.
The equation becomes
\begin{equation}
  k_x^2 + k_z^2 = \epar\,k_0^2.
  \label{eq:ch2_TE_disp}
\end{equation}
Only $\epar$ appears. The out-of-plane permittivity $\eperp$ has completely
dropped out. This happens because the TE electric field is tangential to the
layer planes, so it never probes the out-of-plane response.
Equation~\eqref{eq:ch2_TE_disp} is always the equation of a circle (in 2D)
or sphere (in 3D) of radius $\sqrt{\epar}\,k_0$. It can never be a hyperbola.

There are three progressively deeper reasons why TE hyperbolicity is
impossible, and it is worth stating all three because they illuminate the
physics at different levels.

\paragraph{Reason 1: Algebraic.} The two wavevector components $k_x$ and
$k_z$ enter equation~\eqref{eq:ch2_TE_disp} with the same coefficient. They
can never have opposite signs in the dispersion equation, which is the
algebraic condition for a hyperbola.

\paragraph{Reason 2: Duality.} Restoring the magnetic permeability, the TE
and TM dispersion relations are exact electromagnetic duals of each other.
The TM relation requires $\epar\eperp < 0$ for hyperbolicity. The TE relation
is obtained by the substitution $\eps \leftrightarrow \mu$, so TE
hyperbolicity would require $\mu_\parallel\mu_\perp < 0$: a magnetically
hyperbolic medium. Ni and TiO$_2$ are both non-magnetic at optical frequencies
($\mu_r = 1$), so the TE surface remains a circle.

\paragraph{Reason 3: Physical.} A hyperbolic medium supports propagating
waves with arbitrarily large real wavevectors. For TM, the normal component
$k_z$ remains real even as $k_x \to \infty$ (see equation~\eqref{eq:ch2_kz_aniso}).
For TE, equation~\eqref{eq:ch2_TE_disp} gives $k_z^2 = \epar k_0^2 - k_x^2$,
which becomes large and negative as $k_x$ grows. A negative $k_z^2$ means
$k_z = \pm\ii\kappa_z$ is imaginary, so the field varies as $\ee^{\pm\kappa_z z}$.
In a passive medium, the field must decay away from its source, selecting the
decaying root $\ee^{-\kappa_z z}$. The TE high-$k_x$ mode is therefore
evanescent: it decays exponentially and cannot propagate into the bulk.
Only the TM polarisation supports propagating high-$k$ modes.

\begin{resultbox}
\textbf{Summary: hyperbolicity is a TM-only phenomenon.} In a non-magnetic
uniaxial medium, the TE dispersion is always $k_x^2 + k_z^2 = \epar k_0^2$,
an isotropic circle. Only the TM dispersion
$k_x^2/\eperp + k_z^2/\epar = k_0^2$ involves both permittivity components
and can produce a hyperbolic isofrequency contour. All the interesting physics
of hyperbolic metamaterials is therefore TM physics. The ellipsometric
measurements in this report, which separate the $p$- and $s$-polarised
responses, exploit exactly this distinction.
\end{resultbox}

% -----------------------------------------------------------------------------
\section{The Lossless Isofrequency Contour and its Four Regimes}
% -----------------------------------------------------------------------------

In the idealised limit of a lossless material, both $\epar$ and $\eperp$ are
real. The TM dispersion relation~\eqref{eq:ch2_TM_disp} then defines a conic
section in the $(k_x, k_z)$ plane whose shape depends on the signs of the
two permittivities. There are four cases, summarised in
Table~\ref{tab:ch2_ifc_regimes} and illustrated in
Figure~\ref{fig:ch1_ifc_regimes} in Chapter~\ref{ch:introduction}.

\begin{table}[h]
\centering
\caption{The four lossless TM isofrequency contour topologies, classified
  by the signs of $\Re(\epar)$ and $\Re(\eperp)$. The Ni/TiO$_2$ stack sits
  nominally in the Type-II column, but material losses close the open contour.}
\label{tab:ch2_ifc_regimes}
\begin{tabular}{cclp{5cm}}
\toprule
$\Re(\epar)$ & $\Re(\eperp)$ & IFC shape & Physical character \\
\midrule
$+$ & $+$ & Closed ellipse   & Dielectric in both directions; bounded
                                wavevectors \\
$+$ & $-$ & Type-I hyperbola & Arms along $k_x$; metallic out-of-plane \\
$-$ & $+$ & Type-II hyperbola & Arms along $k_z$; metallic in-plane \\
$-$ & $-$ & No propagating TM modes & Both components metallic \\
\bottomrule
\end{tabular}
\end{table}

For the Ni/TiO$_2$ multilayer, the effective medium theory (Chapter~\ref{ch:emt})
predicts $\Re(\epar) < 0$ at wavelengths above approximately $607$\,nm and
$\Re(\eperp) > 0$ throughout, placing the stack nominally in the Type-II
hyperbolic column. An ideal Type-II hyperboloid opens along the $k_z$ axis with
asymptotes satisfying $k_z/k_x = \pm\sqrt{\eperp/|\epar|}$. The wavevector
magnitude grows without bound as the propagation direction approaches the
asymptote.

In an ideal lossless hyperbolic medium, this unbounded wavevector is a genuine
physical property. It implies that the medium can channel electromagnetic energy
at spatial scales far below one wavelength and that the density of photon states
diverges. These properties have been exploited for near-field heat transfer,
subdiffraction imaging, and enhanced spontaneous emission in low-loss hyperbolic
materials such as hexagonal boron nitride. The question for Ni/TiO$_2$ is
whether the heavy absorption of the Ni layers allows any of these properties
to survive.

% -----------------------------------------------------------------------------
\section{Introducing Loss: Complex Wavevectors and the Collinearity Condition}
% -----------------------------------------------------------------------------

In a real material, the permittivities are complex. We write
\begin{equation}
  \epar = |\epar|\,\ee^{\ii\gamma_\parallel}, \qquad
  \eperp = |\eperp|\,\ee^{\ii\gamma_\perp},
  \label{eq:ch2_polar_eps}
\end{equation}
where $\gamma_\parallel = \arctan(\Im\epar/\Re\epar) \in [0,\pi]$ is the
phase angle of the in-plane permittivity, and similarly for $\gamma_\perp$.
For a dielectric with $\Re\eps > 0$, the phase angle is close to zero.
For a metal with $\Re\eps < 0$ and small loss, the phase angle is close to $\pi$.
For nickel, where $\Im\eps \approx |\Re\eps|$, the phase angle is near $\pi/2$.

With complex permittivities, equation~\eqref{eq:ch2_TM_disp} cannot be
satisfied by a real wavevector. The wavevector must itself become complex:
\begin{equation}
  \kvec = \kvec' + \ii\kvec'',
  \label{eq:ch2_complex_k}
\end{equation}
where $\kvec' = \Re(\kvec)$ is the phase (or propagation) vector and
$\kvec'' = \Im(\kvec)$ is the attenuation vector. The wave amplitude varies as
$\ee^{-\kvec''\cdot\bm{r}}$, which decays along the direction of $\kvec''$.

For a well-behaved plane wave that does not grow transversely (which would
violate energy conservation in a passive medium), the planes of constant phase
must be parallel to the planes of constant amplitude. This means $\kvec'$
and $\kvec''$ must point in the same direction. Writing each complex component
as $k_j = |k_j|\ee^{\ii\phi_j}$ for $j = x, z$, the condition
$\kvec'\parallel\kvec''$ requires the phases to be equal:
\begin{equation}
  \phi_z = \phi_x \equiv \phi \quad \text{(collinearity condition).}
  \label{eq:ch2_collinearity}
\end{equation}
This single constraint, imposed by the requirement of a physical plane wave,
is the key that unlocks the analytical solution. It allows us to write the
entire complex wavevector as a single complex prefactor times a real direction:
\begin{equation}
  \frac{\kvec}{k_0} = \ee^{\ii\phi}\,f(\theta)
  \begin{pmatrix}\sin\theta \\ \cos\theta\end{pmatrix},
  \label{eq:ch2_k_parametric}
\end{equation}
where $\theta$ is a real propagation angle (the direction of energy flow),
$f(\theta) > 0$ is a real magnitude, and $\phi$ is the shared phase. The
complexity of $\kvec$ resides entirely in the prefactor $\ee^{\ii\phi}$.
Both $\theta$ and $\phi$ are real; sweeping $\theta$ from $0$ to $90^\circ$
traces out the full isofrequency surface.

% -----------------------------------------------------------------------------
\section{Solving the Complex Dispersion: Phase and Magnitude Conditions}
% -----------------------------------------------------------------------------

Substituting~\eqref{eq:ch2_k_parametric} into the TM dispersion
relation~\eqref{eq:ch2_TM_disp} gives
\begin{equation}
  1 = \ee^{2\ii\phi}\,f^2\,S(\theta), \qquad
  S(\theta) \equiv \frac{\sin^2\theta}{\eperp} + \frac{\cos^2\theta}{\epar}.
  \label{eq:ch2_S_def}
\end{equation}
Writing $1/\eperp = |\eperp|^{-1}\ee^{-\ii\gamma_\perp}$ and
$1/\epar = |\epar|^{-1}\ee^{-\ii\gamma_\parallel}$, the quantity $S$ is
a complex number: $S = |S|\,\ee^{\ii\beta}$. The left-hand side is the real
number 1, so equating magnitudes and phases of both sides
splits~\eqref{eq:ch2_S_def} into two separate conditions.

\paragraph{Phase condition (determines $\phi$).}
Setting the total phase to zero ($2\phi + \beta = 0$) and expanding
$\beta = \arctan(\Im S / \Re S)$, where
\begin{align}
  \Re S &= \frac{\sin^2\theta}{|\eperp|}\cos\gamma_\perp
         + \frac{\cos^2\theta}{|\epar|}\cos\gamma_\parallel, \\
  \Im S &= -\left(\frac{\sin^2\theta}{|\eperp|}\sin\gamma_\perp
         + \frac{\cos^2\theta}{|\epar|}\sin\gamma_\parallel\right),
\end{align}
gives the shared phase:
\begin{equation}
  \tan(2\phi) = \frac{|\epar|\sin\gamma_\perp\sin^2\theta
                    + |\eperp|\sin\gamma_\parallel\cos^2\theta}
                     {|\epar|\cos\gamma_\perp\sin^2\theta
                    + |\eperp|\cos\gamma_\parallel\cos^2\theta}.
  \label{eq:ch2_phase_condition}
\end{equation}
This is equation (2b) of Jackson et al.~\cite{jackson2024} (with $\mu_r = 1$).
Note that $\phi$ depends on $\theta$: as the propagation direction rotates
from $0^\circ$ (along the optic axis) to $90^\circ$ (perpendicular to it),
$\phi$ changes continuously from $\gamma_\parallel/2$ to $\gamma_\perp/2$.

\paragraph{Magnitude condition (determines $f$).}
From $f^2|S| = 1$ we get $f = |S|^{-1/2}$. Writing $a = \sin^2\theta/|\eperp|$
and $b = \cos^2\theta/|\epar|$:
\begin{equation}
  f(\theta) = \left(\frac{\sin^4\theta}{|\eperp|^2}
            + \frac{\cos^4\theta}{|\epar|^2}
            + \frac{2\cos(\gamma_\perp - \gamma_\parallel)\sin^2\theta\cos^2\theta}
                   {|\eperp||\epar|}\right)^{-1/4}.
  \label{eq:ch2_magnitude_condition}
\end{equation}
This is equation (2c) of Jackson et al.~\cite{jackson2024}.

Equations~\eqref{eq:ch2_phase_condition} and~\eqref{eq:ch2_magnitude_condition}
together with~\eqref{eq:ch2_k_parametric} constitute the complete analytical
solution to the lossy dispersion problem. Separating
$\kvec/k_0 = f\ee^{\ii\phi}(\sin\theta, \cos\theta)$ into real and imaginary
parts:
\begin{align}
  \frac{\Re(\kvec)}{k_0} &= f(\theta)\cos\phi(\theta)
    \begin{pmatrix}\sin\theta\\\cos\theta\end{pmatrix},
  \label{eq:ch2_Rek} \\
  \frac{\Im(\kvec)}{k_0} &= f(\theta)\sin\phi(\theta)
    \begin{pmatrix}\sin\theta\\\cos\theta\end{pmatrix}.
  \label{eq:ch2_Imk}
\end{align}
Sweeping $\theta$ from $0^\circ$ to $90^\circ$ traces the phase surface
(IFC) from~\eqref{eq:ch2_Rek} and the attenuation surface
from~\eqref{eq:ch2_Imk}. The ratio $\Re(k)/\Im(k) = \cot\phi$ tells us
directly how many wavelengths a mode travels before its amplitude decays
by a factor of $1/\ee$.

% -----------------------------------------------------------------------------
\section{Why Loss Closes the Hyperboloid: the Maximum Wavevector}
% -----------------------------------------------------------------------------

We can now understand analytically why material loss closes the open
hyperbola into a finite contour. This is the central physical insight of
the Jackson et al.\ (2024) paper~\cite{jackson2024}.

In the lossless limit, $\gamma_\parallel \to \pi$ (metallic, exactly) and
$\gamma_\perp \to 0$ (dielectric, exactly). The phase condition gives
$\phi \to 0$, so $\ee^{\ii\phi} \to 1$ and $\Re(\kvec) = k_0 f(\theta)(\sin\theta,
\cos\theta)$. As $\theta$ approaches the asymptotic angle of the hyperbola,
the denominator in equation~\eqref{eq:ch2_magnitude_condition} approaches
zero and $f(\theta) \to \infty$. The lossless real wavevector grows without
bound, and the IFC is open.

In the lossy case, $\phi$ is not zero. As $\theta$ increases from $0$ toward
$90^\circ$, the phase $\phi$ varies continuously from $\gamma_\parallel/2$
to $\gamma_\perp/2$. For Ni, $\gamma_\parallel \approx \pi - \arctan(20/18)
\approx 0.83\pi$ (large loss), so $\phi$ starts near $0.42\pi$, which means
$\cos\phi \approx \cos(0.42\pi) \approx 0.25$. The magnitude of the
\emph{real} part of $\kvec$ is
\begin{equation}
  |\Re(\kvec)| = k_0\,f(\theta)\cos\phi(\theta).
  \label{eq:ch2_Rk_product}
\end{equation}
As $\theta$ sweeps toward the old asymptotic direction, $f(\theta)$ grows
but $\cos\phi(\theta)$ falls simultaneously. Their product reaches a finite
maximum and then turns back. The formerly infinite arms of the hyperbola
become closed, and the IFC is a bounded curve.

The maximum wavevector $k_m$ occurs at the propagation angle $\theta_{k_m}$
where $\partial|\Re(\kvec)|/\partial\theta = 0$, which corresponds to a
shared phase
\begin{equation}
  \phi_{k_m} = \frac{\pi}{6} + \frac{1}{3}\arctan\!\left(
    \frac{|\epar|\sin\gamma_\perp - |\eperp|\sin\gamma_\parallel}
         {|\epar|\cos\gamma_\perp - |\eperp|\cos\gamma_\parallel}\right).
  \label{eq:ch2_phikm}
\end{equation}
For small loss, $\phi_{k_m} \approx \pi/6$, giving $k_m \approx
(\sqrt{3}/2)\,k_0 f(\theta_{k_m})$.

Figure~\ref{fig:ch2_lossy_ifc} illustrates this transformation schematically.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.0,
  axis/.style={-{Latex[length=2mm]}, thick, color=navy},
  lossy/.style={very thick, color=KITgreen},
  lossless/.style={very thick, dashed, color=coral},
  attn/.style={thick, color=KITblue, densely dotted},
  lbl/.style={font=\small}
]

%% --- Left panel: lossless ---
\begin{scope}[shift={(-3.8,0)}]
  \draw[axis] (-0.5,0) -- (2.2,0) node[right, lbl] {$k_x/k_0$};
  \draw[axis] (0,-2.0) -- (0,2.0) node[above, lbl] {$k_z/k_0$};

  % Type-II hyperbola (arms along kz); domain >= sqrt(0.18) ~ 0.424, use 0.44
  \foreach \sgn in {1,-1}{
    \draw[lossless, domain=0.44:1.9, samples=80]
      plot ({0.55*sqrt(\x*\x - 0.18)}, {\sgn*\x});
  }

  % Asymptote
  \draw[gray!50, dashed] (0,-1.9) -- (1.05, 1.9);
  \draw[gray!50, dashed] (0, 1.9) -- (1.05,-1.9);
  \node[below, font=\footnotesize, color=coral] at (1.1,-2.1)
    {Lossless (Type-II)};
  \node[font=\scriptsize, color=gray] at (1.6,1.5) {asymptote};

  \draw[-{Latex[length=2mm]}, gray, thick] (0.5, 1.5) -- (0.7, 1.85);
  \node[right, font=\scriptsize, color=gray] at (0.7, 1.7) {$k\to\infty$};
\end{scope}

%% --- Right panel: lossy ---
\begin{scope}[shift={(2.5,0)}]
  \draw[axis] (-0.5,0) -- (2.2,0) node[right, lbl] {$k_x/k_0$};
  \draw[axis] (0,-2.0) -- (0,2.0) node[above, lbl] {$k_z/k_0$};

  % Re(k): closed lossy IFC
  \draw[lossy] (0,0)
    .. controls (0.25, 1.1) and (0.8, 1.4) .. (1.35, 0.65)
    .. controls (1.65, 0.15) and (1.65,-0.15) .. (1.35,-0.65)
    .. controls (0.8,-1.4) and (0.25,-1.1) .. (0,0);

  % km marker
  \fill[KITgreen] (1.65, 0) circle (2pt);
  \draw[dashed, KITgreen!60] (1.65,0) -- (1.65,-0.35);
  \node[right, font=\scriptsize, color=KITgreen] at (1.65,-0.35)
    {$k_m$};

  % Im(k): attenuation surface
  \draw[attn] (0,0)
    .. controls (0.15, 0.7) and (0.55, 1.1) .. (1.0, 1.2)
    .. controls (1.35, 1.3) and (1.7, 1.1) .. (1.9, 0.65)
    .. controls (2.0, 0.3) and (2.0,-0.3) .. (1.9,-0.65)
    .. controls (1.7,-1.1) and (1.35,-1.3) .. (1.0,-1.2)
    .. controls (0.55,-1.1) and (0.15,-0.7) .. (0,0);

  \node[below, font=\footnotesize, color=KITgreen] at (1.1,-2.1)
    {Lossy (closed)};
  \node[right, font=\scriptsize, color=KITgreen] at (0.5, 1.3)
    {$\Re(\kvec)/k_0$};
  \node[right, font=\scriptsize, color=KITblue] at (1.7, 1.1)
    {$\Im(\kvec)/k_0$};
\end{scope}

\draw[-{Latex[length=3mm]}, very thick, color=navy!60]
  (-1.4, 0) -- (-0.2, 0)
  node[midway, above, font=\footnotesize, color=navy] {add loss};

\end{tikzpicture}
\caption{Schematic of how material loss transforms the isofrequency contour
  (IFC) of a Type-II hyperbolic medium.
  \textbf{Left:} the lossless case. The IFC is an open hyperbola whose arms
  extend to arbitrarily large wavevectors along the $k_z$ axis. Grey dashed
  lines show the asymptotes.
  \textbf{Right:} with complex permittivities. The phase $\phi(\theta)$
  introduced by loss causes $\cos\phi$ to suppress $|\Re(\kvec)|$ as the
  wavenumber grows, closing the arms at a finite maximum wavevector $k_m$
  (filled circle). The attenuation surface $\Im(\kvec)/k_0$ (dotted blue)
  continues to grow beyond $k_m$: the largest-wavevector modes are also
  the most strongly damped.}
\label{fig:ch2_lossy_ifc}
\end{figure}

The key physical implication is that $k_m$ sets an upper limit on the
spatial resolution that a hyperbolic metamaterial can provide. Modes with
$k > k_m$ do not exist as propagating waves; they are purely evanescent.
The figure of merit of a lossy hyperbolic medium is therefore $k_m/k_0$.

% -----------------------------------------------------------------------------
\section{Application to the Ni/TiO\texorpdfstring{$_2$}{2} System}
% -----------------------------------------------------------------------------

We can now evaluate the Ni/TiO$_2$ stack using the framework developed above.
The key dimensionless quantity is the \textbf{in-plane loss tangent}
$\Im(\epar)/|\Re(\epar)|$. A hyperbolic medium with open propagating branches
requires this ratio to be much less than 1. When the loss tangent is of order
1 or larger, the IFC is closed and the hyperbolic character is lost.

For Ni/TiO$_2$ with fill fraction $\rho \approx 0.379$, using the Johnson
and Christy Ni data~\cite{johnson_christy} and Jolivet TiO$_2$
data~\cite{jolivet}:

\begin{center}
\begin{tabular}{cccc}
\toprule
$\lambda$ (nm) & $\Re(\epar)$ & $\Im(\epar)$ &
  $\Im(\epar)/|\Re(\epar)|$ \\
\midrule
600  & $+0.11$ & $5.15$ & $\gg 1$ (near ENZ) \\
650  & $-0.73$ & $5.58$ & $7.6$  \\
700  & $-1.47$ & $6.13$ & $4.2$  \\
750  & $-2.10$ & $6.65$ & $3.2$  \\
850  & $-3.00$ & $7.40$ & $2.5$  \\
1000 & $-4.00$ & $8.00$ & $2.0$  \\
\bottomrule
\end{tabular}
\end{center}

The loss tangent ranges from 2 to above 7 across the measurement window.
It is far above 1 throughout. In a hyperbolic material like hexagonal boron
nitride (hBN), by contrast, the analogous loss tangent is of order $10^{-2}$.
Ni/TiO$_2$ is a \textbf{heavily lossy} system in which the imaginary part of
$\epar$ dominates the real part at every wavelength studied.

Using the Jackson et al.\ (2024) framework, the maximum wavevector for this
system is approximately $k_m/k_0 \approx 3.1$--$3.2$ across $550$--$1000$\,nm.
For comparison, hBN reaches $k_m/k_0 \approx 7$--$12$ in its hyperbolic bands.
The Ni/TiO$_2$ contour is nominally Type-II by the lossless classification,
but the dominant imaginary part forces it into a closed shape rather than an
open hyperbola.

The extraction results in Chapter~\ref{ch:results} will further show that
both $\Re(\epar)$ and $\Re(\eperp)$ are positive throughout $550$--$1000$\,nm
once the actual fabricated sample is characterised by ellipsometry. The
effective medium prediction of a sign change in $\Re(\epar)$ near $607$\,nm
is not borne out by the measurement, because the fabricated Ni is sub-metallic
relative to the bulk Johnson and Christy values. The IFC is therefore a closed
ellipse in both the lossless and lossy sense, as quantified in
Chapter~\ref{ch:results}.

\begin{resultbox}
\textbf{Summary of Chapter~\ref{ch:em_theory}.} Starting from Maxwell's
equations, we derived the TM dispersion relation
$k_x^2/\eperp + k_z^2/\epar = k_0^2$, which governs wave propagation in a
uniaxial medium. We showed that the TE polarisation is always isotropic and
can never be hyperbolic in a non-magnetic material. We derived the full
analytical solution for the lossy case following Jackson et al.\
(2024)~\cite{jackson2024}, showing that material loss forces the wavevector
to be complex, requires a collinearity condition, and produces a phase factor
$\ee^{\ii\phi}$ that suppresses $|\Re(\kvec)|$ at large $\theta$, closing
the formerly open hyperboloid at a finite maximum wavevector $k_m$. For
Ni/TiO$_2$ with in-plane loss tangent $\sim 2$--$8$, the IFC is a closed
loss-broadened contour throughout the measurement window.
\end{resultbox}
