% =============================================================================
% Chapter 5: Transfer Matrix Method -- Step-by-Step Derivation
% =============================================================================

\chapter{Transfer Matrix Method: Step-by-Step Derivation}
\label{ch:tmm_derivation}

A multilayer optical stack presents a cascade of boundary matching problems.
At every interface between two materials, the electromagnetic fields on both
sides must be consistent with Maxwell's equations. For a stack with $N$ layers
and therefore $N+1$ interfaces, a brute-force approach would require solving
$2(N+1)$ simultaneous linear equations for the unknown field amplitudes. For
our six-layer Ni/TiO$_2$ stack this is manageable, but the algebra quickly
becomes unwieldy for more complex structures, and there is no systematic way
to see how the result depends on the individual layer parameters.

The \textbf{transfer matrix method} (TMM) solves this problem by encoding the
field matching at each interface and the propagation through each layer as a
series of $2\times 2$ matrix multiplications. The total optical response of the
stack is then the product of these matrices. Adding or removing a layer changes
only one factor in the product, and the physical effect of each layer is
immediately visible in its matrix. This chapter derives the method from
scratch, step by step, and applies it to the six-layer Ni/TiO$_2$ stack using
the exact FIB/TEM-measured thicknesses.

% -----------------------------------------------------------------------------
\section{Fields in a Homogeneous Layer}
% -----------------------------------------------------------------------------

We work in the geometry established in Chapter~\ref{ch:em_theory}: the stack
lies in the $xy$ plane, the layer normals point along $z$, and the plane of
incidence is the $xz$ plane. We consider a monochromatic plane wave of angular
frequency $\omega$ with the time convention $\ee^{-\ii\omega t}$.

The in-plane wavevector component $k_x = k_0\sin\theta_\text{inc}$ is
conserved across all interfaces (equation~\eqref{eq:ch2_kx_conserved}), where
$\theta_\text{inc}$ is the angle of incidence in the ambient medium. Inside any
layer $j$ with complex permittivity $\tilde\eps_j$, the normal component of the
wavevector is
\begin{equation}
  k_{z,j} = \sqrt{\tilde\eps_j\,k_0^2 - k_x^2}.
  \label{eq:ch5_kz_isotropic}
\end{equation}
We choose the root with $\Im(k_{z,j}) > 0$ for absorbing media, so that the
downward-propagating wave $\ee^{+\ii k_{z,j} z}$ decays in amplitude rather than
growing without bound as $z$ increases.

Inside layer $j$, the general field is a superposition of a downward-propagating
wave (amplitude $A_j$) and an upward-propagating wave (amplitude $B_j$). For
the TE (s-polarised) polarisation, the electric field is:
\begin{equation}
  E_y^{(j)}(z) = A_j\,\ee^{+\ii k_{z,j} z} + B_j\,\ee^{-\ii k_{z,j} z},
  \label{eq:ch5_Ey}
\end{equation}
and the corresponding magnetic field component (from Faraday's law) is:
\begin{equation}
  H_x^{(j)}(z) = \frac{k_{z,j}}{k_0}\!\left(
    A_j\,\ee^{+\ii k_{z,j} z} - B_j\,\ee^{-\ii k_{z,j} z}
  \right),
  \label{eq:ch5_Hx}
\end{equation}
where factors of $\eps_0 c$ have been absorbed into the field amplitude.
We define the TE optical admittance
\begin{equation}
  \eta_j^{(s)} = \frac{k_{z,j}}{k_0},
  \label{eq:ch5_eta_s}
\end{equation}
so that $H_x^{(j)} = \eta_j^{(s)}(A_j\,\ee^{+\ii k_{z,j} z} -
B_j\,\ee^{-\ii k_{z,j} z})$. The field vector at position $z$ inside layer $j$
can then be written as
\begin{equation}
  \begin{pmatrix} E_y \\ H_x \end{pmatrix}_z
  = \begin{pmatrix} 1 & 1 \\ \eta^{(s)} & -\eta^{(s)} \end{pmatrix}
    \begin{pmatrix} A_j\,\ee^{+\ii k_{z,j} z} \\ B_j\,\ee^{-\ii k_{z,j} z} \end{pmatrix}.
  \label{eq:ch5_field_vec}
\end{equation}

For the TM (p-polarised) polarisation, the roles of $E$ and $H$ are exchanged.
The tangential fields are $H_y$ and $E_x$, and the TM optical admittance is
\begin{equation}
  \eta_j^{(p)} = \frac{\tilde\eps_j\,k_0}{k_{z,j}}.
  \label{eq:ch5_eta_p}
\end{equation}
The structure of the field vectors is identical to the TE case with
$\eta^{(s)} \to \eta^{(p)}$ and $E_y \to H_y$, $H_x \to E_x$. All the
derivations below hold for both polarisations by using the appropriate
admittance $\eta$.

% -----------------------------------------------------------------------------
\section{Boundary Conditions at a Single Interface}
% -----------------------------------------------------------------------------

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.1,
  ifc/.style={very thick, color=navy},
  arr/.style={-{Latex[length=2mm]}, thick},
  lbl/.style={font=\small}
]
  % Regions
  \fill[KITblue!10] (-3.5, 0) rectangle (3.5, 2.2);
  \fill[gray!20]    (-3.5,-2.2) rectangle (3.5, 0);
  \draw[ifc] (-3.5,0) -- (3.5,0);

  % Labels
  \node[lbl] at (-2.5, 1.4) {Layer $j$ (above)};
  \node[lbl, color=navy!70] at (-2.5, 0.7) {$\tilde\eps_j,\; k_{z,j},\; \eta_j$};
  \node[lbl] at (-2.5,-1.4) {Layer $j{+}1$ (below)};
  \node[lbl, color=navy!70] at (-2.5,-0.7) {$\tilde\eps_{j+1},\; k_{z,j+1},\; \eta_{j+1}$};
  \node[right, lbl, color=navy] at (3.5, 0.0) {$z = z_j$};

  % Wave arrows (layer j)
  \draw[arr, color=KITblue, very thick] (0.5, 2.0) -- (0.5, 0.2)
    node[midway, right, lbl] {$A_j\,\ee^{+\ii k_{z,j}z}$};
  \draw[arr, color=coral, very thick] (1.5, 0.2) -- (1.5, 2.0)
    node[midway, right, lbl] {$B_j\,\ee^{-\ii k_{z,j}z}$};

  % Wave arrows (layer j+1)
  \draw[arr, color=KITblue, very thick] (0.5,-0.2) -- (0.5,-2.0)
    node[midway, right, lbl] {$A_{j+1}\,\ee^{+\ii k_{z,j+1}z}$};
  \draw[arr, color=coral, very thick] (1.5,-2.0) -- (1.5,-0.2)
    node[midway, right, lbl] {$B_{j+1}\,\ee^{-\ii k_{z,j+1}z}$};

  % Boundary condition note
  \node[left, align=right, font=\footnotesize, color=navy] at (-0.2, 0) {
    $E_y$ (or $H_y$) continuous\\
    $H_x$ (or $E_x$) continuous};
\end{tikzpicture}
\caption{Field amplitudes at the interface between layer $j$ (above) and
  layer $j+1$ (below) at position $z = z_j$. Each layer supports a
  downward-propagating wave (amplitude $A$, blue arrow) and an
  upward-propagating wave (amplitude $B$, red arrow). The boundary conditions
  require the tangential field components to be the same on both sides of the
  interface.}
\label{fig:ch5_interface}
\end{figure}

At the interface $z = z_j$ between layer $j$ and layer $j+1$, the boundary
conditions require the tangential components of $\Evec$ and $\Hvec$ to be
continuous. Evaluating equation~\eqref{eq:ch5_field_vec} on both sides at
$z = z_j$ and setting the fields equal:
\begin{equation}
  \begin{pmatrix} 1 & 1 \\ \eta_j & -\eta_j \end{pmatrix}
  \begin{pmatrix} A_j\,\ee^{+\ii k_{z,j} z_j} \\ B_j\,\ee^{-\ii k_{z,j} z_j} \end{pmatrix}
  =
  \begin{pmatrix} 1 & 1 \\ \eta_{j+1} & -\eta_{j+1} \end{pmatrix}
  \begin{pmatrix} A_{j+1}\,\ee^{+\ii k_{z,j+1} z_j} \\
                  B_{j+1}\,\ee^{-\ii k_{z,j+1} z_j} \end{pmatrix}.
  \label{eq:ch5_bc_raw}
\end{equation}
We call the $2\times 2$ matrix on each side the \textbf{dynamical matrix}
$\bm{D}_j$:
\begin{equation}
  \bm{D}_j = \begin{pmatrix} 1 & 1 \\ \eta_j & -\eta_j \end{pmatrix}.
  \label{eq:ch5_dynamical}
\end{equation}
Inverting $\bm{D}_j$ (which has determinant $-2\eta_j$):
\begin{equation}
  \bm{D}_j^{-1} = \frac{1}{2}
  \begin{pmatrix} 1 & \eta_j^{-1} \\ 1 & -\eta_j^{-1} \end{pmatrix},
  \label{eq:ch5_dinv}
\end{equation}
the field amplitude vector $(A_j, B_j)^T$ at the top of the interface relates
to the vector at the bottom by:
\begin{equation}
  \begin{pmatrix} A_j \\ B_j \end{pmatrix}_{z_j}
  = \bm{D}_j^{-1}\,\bm{D}_{j+1}
  \begin{pmatrix} A_{j+1} \\ B_{j+1} \end{pmatrix}_{z_j}.
  \label{eq:ch5_transfer_interface}
\end{equation}
The matrix $\bm{D}_j^{-1}\bm{D}_{j+1}$ is the \textbf{interface transfer
matrix}.

% -----------------------------------------------------------------------------
\section{Propagation Through a Single Layer}
% -----------------------------------------------------------------------------

Inside layer $j$, the wave amplitudes at position $z$ are those at $z=0$
multiplied by phase factors:
\begin{equation}
  A_j(z) = A_j(0)\,\ee^{+\ii k_{z,j} z}, \qquad
  B_j(z) = B_j(0)\,\ee^{-\ii k_{z,j} z}.
  \label{eq:ch5_phase_factors}
\end{equation}
For a layer of thickness $d_j$ extending from $z=0$ to $z=d_j$, the amplitudes
at the bottom relate to those at the top by the \textbf{propagation matrix}:
\begin{equation}
  \begin{pmatrix} A_j \\ B_j \end{pmatrix}_{z=d_j}
  = \begin{pmatrix} \ee^{+\ii k_{z,j} d_j} & 0 \\ 0 & \ee^{-\ii k_{z,j} d_j}
    \end{pmatrix}
  \begin{pmatrix} A_j \\ B_j \end{pmatrix}_{z=0}.
  \label{eq:ch5_prop_matrix}
\end{equation}

% -----------------------------------------------------------------------------
\section{The Abelès Characteristic Matrix}
% -----------------------------------------------------------------------------

Combining the propagation through layer $j$ and the interface matching, the
field vector at the top of layer $j$ relates to the field vector at the bottom
through the \textbf{Abelès characteristic matrix}~\cite{macleod}:
\begin{equation}
  \bm{M}_j = \begin{pmatrix}
    \cos\delta_j & -\ii\,\sin\delta_j / \eta_j \\
    -\ii\,\eta_j\,\sin\delta_j & \cos\delta_j
  \end{pmatrix},
  \label{eq:ch5_abeles}
\end{equation}
where the \textbf{phase thickness} is
\begin{equation}
  \boxed{\delta_j = -k_{z,j}\,d_j.}
  \label{eq:ch5_delta}
\end{equation}

\begin{warnbox}
\textbf{Sign convention for $\delta$.} With the time convention
$\ee^{-\ii\omega t}$, the downward-propagating wave carries the factor
$\ee^{+\ii k_{z,j} z}$. For the Abelès matrix to take the cosine-sine form in
equation~\eqref{eq:ch5_abeles}, the phase variable must be defined as
$\delta_j = -k_{z,j} d_j$. Using the opposite sign $\delta_j = +k_{z,j} d_j$
is a common error that produces results disagreeing with measured data and
with the \texttt{tmm} Python package. The correct sign gives agreement to
$10^{-14}$ degrees with the \texttt{tmm} package in all validation tests.
\end{warnbox}

% -----------------------------------------------------------------------------
\section{The Full Multilayer: Ordered Matrix Product}
% -----------------------------------------------------------------------------

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.9,
  box/.style={draw=navy, fill=navy!8, minimum width=1.8cm,
              minimum height=0.9cm, font=\small, align=center},
  amb/.style={draw=KITgreen, fill=KITgreen!10, minimum width=1.2cm,
              minimum height=0.6cm, font=\small, align=center},
  arr/.style={-{Latex[length=2mm]}, thick, color=navy}
]
\node[amb] (amb) at (-5.5, 0) {Ambient\\($\eta_0$)};
\node[box] (m1) at (-3.3, 0) {$\bm{M}_1$\\(TiO$_2$, $d_1$)};
\node[box] (m2) at (-1.1, 0) {$\bm{M}_2$\\(Ni, $d_2$)};
\node[font=\Large] at (0.5, 0) {$\cdots$};
\node[box] (m6) at (2.1, 0) {$\bm{M}_6$\\(Ni, $d_6$)};
\node[amb] (sub) at (4.0, 0) {Substrate\\($\eta_s$)};
\draw[arr] (amb.east) -- (m1.west);
\draw[arr] (m1.east) -- (m2.west);
\draw[arr] (m2.east) -- (0.0, 0);
\draw[arr] (1.0, 0) -- (m6.west);
\draw[arr] (m6.east) -- (sub.west);
\node[above=10pt, font=\small, color=KITblue] at ($(m1.north)!0.5!(m6.north)$)
  {$\bm{M}_\text{tot} = \bm{M}_1\cdot\bm{M}_2\cdots\bm{M}_6$};
\draw[color=KITblue, thick]
  (m1.north west) -- ++(0, 6pt) -- ($(m6.north east)+(0,6pt)$) -- (m6.north east);
\draw[arr, color=navy!40] (-6.5,-0.8) -- (5.2,-0.8)
  node[right, font=\scriptsize] {$z$};
\node[below, font=\scriptsize, color=navy!40] at (-5.5,-0.8) {air};
\node[below, font=\scriptsize, color=navy!40] at (4.0,-0.8) {SiO$_2$};
\end{tikzpicture}
\caption{Matrix product structure for the six-layer [TiO$_2$/Ni]$\times$3
  stack. The total optical response is the ordered product
  $\bm{M}_\text{tot} = \bm{M}_1 \cdot \bm{M}_2 \cdots \bm{M}_6$, computed
  from the incident-medium side (left) to the substrate side (right). The
  ambient medium and substrate enter only through their admittances in the
  B/C vector calculation.}
\label{fig:ch5_matrix_product}
\end{figure}

For a stack of $N$ layers, the total characteristic matrix is:
\begin{equation}
  \bm{M}_\text{tot} = \prod_{j=1}^{N} \bm{M}_j
  = \begin{pmatrix} m_{11} & m_{12} \\ m_{21} & m_{22} \end{pmatrix}.
  \label{eq:ch5_mtot}
\end{equation}
The order matters: $\bm{M}_1$ is the first layer encountered by the incoming
wave. For our stack with TiO$_2$ facing the air, $\bm{M}_1$ is the first
TiO$_2$ layer.

% -----------------------------------------------------------------------------
\section{Extracting R and T: the B/C Vector Formalism}
% -----------------------------------------------------------------------------

We define the column vector
\begin{equation}
  \begin{pmatrix} B \\ C \end{pmatrix}
  = \bm{M}_\text{tot}
  \begin{pmatrix} 1 \\ \eta_s \end{pmatrix},
  \label{eq:ch5_BC}
\end{equation}
where $\eta_s$ is the admittance of the substrate. The physical content of
this operation is: we seed the substrate side with the field of a transmitted
wave and propagate it back through all layers to the ambient interface. The
reflection and transmission amplitudes follow:
\begin{align}
  r &= \frac{\eta_0 B - C}{\eta_0 B + C},
  \label{eq:ch5_r} \\
  t &= \frac{2\eta_0}{\eta_0 B + C},
  \label{eq:ch5_t}
\end{align}
where $\eta_0$ is the admittance of the ambient medium. The power reflectance
and transmittance are
\begin{align}
  R &= |r|^2,
  \label{eq:ch5_R} \\
  T &= \frac{\Re(\eta_s)}{\Re(\eta_0)}\,|t|^2.
  \label{eq:ch5_T}
\end{align}
The factor $\Re(\eta_s)/\Re(\eta_0)$ accounts for the ratio of real power in
the transmitted and incident waves. The absorbed fraction is $A = 1 - R - T$.

\begin{resultbox}
\textbf{Energy conservation check.} For a passive stack, $R + T \leq 1$ at
every wavelength. Any negative absorptance signals a sign error in $\delta$,
a wrong branch choice for $k_z$, or an inconsistency in the admittance
definitions. This check was used throughout the development of the code.
\end{resultbox}

% -----------------------------------------------------------------------------
\section{Extension to Uniaxial Anisotropic Layers}
% -----------------------------------------------------------------------------

The Abelès formalism above applies to isotropic layers. The Ni/TiO$_2$ stack,
treated as a homogeneous effective medium, is uniaxial with in-plane
permittivity $\epar$ and out-of-plane permittivity $\eperp$. Only two
substitutions are needed to extend the formalism.

\subsection{Modified Wavevector for the Extraordinary Wave}

For the \textbf{TE} wave in a uniaxial layer, the electric field is purely
in-plane and probes only $\epar$. The wavevector is identical to the isotropic
case with $\tilde\eps \to \epar$:
\begin{equation}
  k_{z}^{(\mathrm{TE})} = \sqrt{\epar\,k_0^2 - k_x^2},
  \qquad \eta^{(s)} = \frac{k_z^{(\mathrm{TE})}}{k_0}.
  \label{eq:ch5_kz_TE}
\end{equation}

For the \textbf{TM} wave, the electric field has both in-plane and out-of-plane
components. From the TM dispersion relation (equation~\eqref{eq:ch2_TM_disp}):
\begin{equation}
  k_{z}^{(\mathrm{TM})} = k_0\sqrt{\epar - \frac{\epar}{\eperp}\sin^2\theta}.
  \label{eq:ch5_kz_TM}
\end{equation}
This is the \textbf{extraordinary wavevector}. Both $\epar$ and $\eperp$
appear: the TM wave samples both permittivity components simultaneously.
This is the key mathematical reason why TM ellipsometry is sensitive to
$\eperp$ while TE is not.

\subsection{Modified Admittance for TM}

For TM in a uniaxial medium, the tangential fields $H_y$ and $E_x$ are
related through the \textbf{tilted admittance}:
\begin{equation}
  \eta^{(p)}_\text{aniso} = \frac{\epar\,k_0}{k_z^{(\mathrm{TM})}}.
  \label{eq:ch5_eta_p_aniso}
\end{equation}
For an isotropic medium ($\epar = \eperp = \tilde\eps$), this reduces to
$\tilde\eps k_0/k_z$, recovering the standard result.

\subsection{The Anisotropic Characteristic Matrix}

With these two substitutions, the Abelès matrix takes exactly the same form:
\begin{equation}
  \bm{M}_j^\text{aniso} = \begin{pmatrix}
    \cos\delta_j & -\ii\,\sin\delta_j / \eta_j^{(p)} \\
    -\ii\,\eta_j^{(p)}\,\sin\delta_j & \cos\delta_j
  \end{pmatrix},
  \qquad \delta_j = -k_{z,j}^{(\mathrm{TM})}\,d_j.
  \label{eq:ch5_abeles_aniso}
\end{equation}
The entire B/C formalism applies unchanged. This is the key practical
advantage of the Abelès approach: uniaxial anisotropy requires only two
parameter substitutions; the matrix algebra is identical.

% -----------------------------------------------------------------------------
\section{Computing the Ellipsometric Observables}
% -----------------------------------------------------------------------------

We apply the B/C formalism twice: once with TE admittances (yielding $r_s$)
and once with TM admittances (yielding $r_p$):
\begin{align}
  r_s &= \frac{\eta_0^{(s)} B_s - C_s}{\eta_0^{(s)} B_s + C_s},
  \label{eq:ch5_rs} \\
  r_p &= \frac{\eta_0^{(p)} B_p - C_p}{\eta_0^{(p)} B_p + C_p},
  \label{eq:ch5_rp}
\end{align}
where $\eta_0^{(s)} = \cos\theta_\text{inc}$ and
$\eta_0^{(p)} = 1/\cos\theta_\text{inc}$ for air. The ellipsometric ratio
and NCS parameters are then:
\begin{equation}
  \rho_\text{ell} = \frac{r_p}{r_s} = \tan\Psi\,\ee^{\ii\Delta},
  \label{eq:ch5_rho_ell}
\end{equation}
\begin{align}
  N &= \cos 2\Psi = \frac{|r_s|^2 - |r_p|^2}{|r_s|^2 + |r_p|^2},
  \label{eq:ch5_N} \\
  C &= \sin 2\Psi\,\cos\Delta
     = \frac{2\,\Re(r_p r_s^*)}{|r_s|^2 + |r_p|^2},
  \label{eq:ch5_C} \\
  S &= \sin 2\Psi\,\sin\Delta
     = \frac{2\,\Im(r_p r_s^*)}{|r_s|^2 + |r_p|^2}.
  \label{eq:ch5_S}
\end{align}
To match the CompleteEASE convention (Chapter~\ref{ch:sample_measurements}),
$S$ is negated: $S_\text{CE} = -S$.

% -----------------------------------------------------------------------------
\section{Explicit Application to the Ni/TiO\texorpdfstring{$_2$}{2} Stack}
% -----------------------------------------------------------------------------

We apply the full TMM to the [TiO$_2$/Ni]$\times$3 stack at $\lambda = 700$\,nm
and $\theta = 60^\circ$ to give a concrete numerical illustration.

\subsection{Input Parameters at 700\,nm, $60^\circ$}

The free-space wavenumber and in-plane wavevector are:
\begin{align}
  k_0 &= \frac{2\pi}{700\,\text{nm}} = 8.976 \times 10^{-3}\,\text{nm}^{-1}, \\
  k_x &= k_0\sin 60^\circ = 7.772 \times 10^{-3}\,\text{nm}^{-1}.
  \label{eq:ch5_kxk0}
\end{align}
The complex permittivities from the material databases at this wavelength are:
\begin{align}
  \eps_{\mathrm{Ni}}(700\,\text{nm})
    &= (2.12 + 4.70\,\ii)^2 = -17.59 + 19.93\,\ii, \\
  \eps_{\mathrm{TiO_2}}(700\,\text{nm})
    &\approx 5.52\quad\text{(effectively real)}.
  \label{eq:ch5_epsilons}
\end{align}
The wavevectors in each material are:
\begin{align}
  k_{z,\mathrm{TiO_2}} &= k_0\sqrt{5.52 - \sin^2 60^\circ}
                         = k_0 \times 2.184\;\text{(real)},
  \label{eq:ch5_kz_TiO2} \\
  k_{z,\mathrm{Ni}} &= k_0\sqrt{-18.34 + 19.93\,\ii}
                      \approx k_0\,(2.09 + 4.76\,\ii).
  \label{eq:ch5_kz_Ni}
\end{align}

\subsection{Phase Thicknesses and First-Layer Matrix}

For the first TiO$_2$ layer ($d_1 = 9.171$\,nm):
\begin{equation}
  \delta_1 = -k_{z,\mathrm{TiO_2}}\,d_1
           = -2.184 \times 9.171 \times 8.976 \times 10^{-3}
           = -0.1796\;\text{(real)},
  \label{eq:ch5_delta1}
\end{equation}
giving the characteristic matrix (with $\eta_{\mathrm{TiO_2}}^{(s)} = 2.184$):
\begin{equation}
  \bm{M}_1 = \begin{pmatrix}
    \cos(-0.1796) & -\ii\sin(-0.1796)/2.184 \\
    -\ii\times 2.184\times\sin(-0.1796) & \cos(-0.1796)
  \end{pmatrix}
  = \begin{pmatrix}
    0.984 & +\ii\,0.082 \\
    +\ii\,0.394 & 0.984
  \end{pmatrix}.
  \label{eq:ch5_M1_numerical}
\end{equation}
For a lossless layer, $\delta$ is real, the diagonal entries are real, and
the off-diagonal entries are purely imaginary, as expected.

For the first Ni layer ($d_2 = 5.556$\,nm), $\delta_2$ is complex:
\begin{equation}
  \delta_2 = -(2.09 + 4.76\,\ii) \times 5.556 \times 8.976 \times 10^{-3}
           = -0.1042 - 0.2373\,\ii.
  \label{eq:ch5_delta2}
\end{equation}
The matrix entries now involve $\cos(-0.1042 - 0.2373\,\ii)$, computed via
$\cos(\alpha + \ii\beta) = \cos\alpha\cosh\beta - \ii\sin\alpha\sinh\beta$,
giving complex entries with comparable real and imaginary parts. This reflects
the strongly absorbing character of the Ni layer.

The total matrix $\bm{M}_\text{tot} = \bm{M}_1\cdots\bm{M}_6$ is computed
numerically in the code module \texttt{aniso\_tmm.py}, which implements both
the isotropic and anisotropic (uniaxial) cases and is validated to $10^{-14}$
degrees against the \texttt{tmm} Python package.

% -----------------------------------------------------------------------------
\section{Validation}
% -----------------------------------------------------------------------------

The TMM implementation was validated at three levels.

\paragraph{Level 1: Isotropic agreement with \texttt{tmm}.}
For the [TiO$_2$/Ni]$\times$3 stack at all wavelengths ($400$ to $1000$\,nm)
and all angles of incidence ($0$ to $80^\circ$), our implementation agrees
with the \texttt{tmm} package to better than $10^{-14}$ degrees in $\Psi$
and $\Delta$. This exceeds the precision of double-precision arithmetic and
confirms that the matrix algebra and admittance definitions are correct.

\paragraph{Level 2: Anisotropic agreement with Berreman $4\times 4$.}
For uniaxial test cases, the Berreman $4\times 4$ matrix method provides an
independent first-principles reference that makes no assumptions about
polarisation separation. Our $2\times 2$ anisotropic Abelès implementation
agrees with the Berreman method to $10^{-15}$, confirming the extraordinary
$k_z$ and tilted admittance are correctly implemented.

\paragraph{Level 3: Energy conservation.}
The computed $R + T$ was checked against $1 - A$ at every wavelength and
angle. No deviation greater than $10^{-12}$ was found, confirming that the
energy balance is maintained throughout.

\paragraph{Substrate discrimination.}
Running the model with SiO$_2$ (Malitson) vs.~Si (Green 2008) substrates
gave rms ellipsometric residuals of $0.001$ and $0.012$ respectively,
confirming the substrate as borosilicate glass (optically equivalent to
fused silica) and demonstrating that the model is sensitive enough to
discriminate between different substrate materials.

\begin{resultbox}
\textbf{Summary of Chapter~\ref{ch:tmm_derivation}.} Starting from boundary
conditions at a single interface, we derived the Abelès $2\times 2$
characteristic matrix $\bm{M}_j$ with phase thickness $\delta_j = -k_{z,j}d_j$.
The total product $\bm{M}_\text{tot}$ connects to $R$, $T$, and the
ellipsometric observables $N$, $C$, $S$ via the B/C vector formalism.
The uniaxial extension requires only two substitutions: the extraordinary
wavevector $k_z^{(\mathrm{TM})} = k_0\sqrt{\epar - (\epar/\eperp)\sin^2\theta}$
and the tilted admittance $\eta^{(p)} = \epar k_0/k_z^{(\mathrm{TM})}$.
The implementation was validated to $10^{-14}$ degrees against the
\texttt{tmm} package and to $10^{-15}$ against the Berreman $4\times 4$
method.
\end{resultbox}
